Spatial Reasoning and Statistical Graphics

### 2014-11-03 Paper/Draft1/TestingVisualAptitude.Rnw
Remove "visual intelligence" regression
Move lit review summary to below end of document
Remove unnecessary lineup vs. other test plots, comments
Code to fix scatterplot matrix plot names

  rbind.fill(
    data.frame(
      xvar = names(data)[ycol], 
      yvar = names(data)[xcol],
      x = data[, xcol], 
      y = data[, ycol], 
      points = T,
      data
    ),
    data.frame(
      yvar = names(data)[ycol], 
      xvar = names(data)[xcol],
      x = mean(range(data[, ycol])), 
      y = mean(range(data[, xcol])),
      label = round(cor(data[,xcol], data[,ycol]), 3),
      points = F
    )
  )
}))
all$xvar <- factor(all$xvar, levels=names(data))
all$yvar <- factor(all$yvar, levels=names(data))
densities <- do.call("rbind", lapply(1:ncol(data), function(i) {
    tmp <- density(data[,i], cut = 1, adjust=.75)
    data.frame(
      xvar = names(data)[i], 
      yvar = names(data)[i],
      x = tmp$x, 
      y = tmp$y/max(tmp$y)* diff(range(tmp$x)) + min(tmp$x)
    )
  }))

@@

\newtext{This emphasizes the underpinnings of lineups: they are a visual medium, but they ultimately are a classification task, presented in a graphical manner. Using lineups as a proxy for statistical significance tests is similar to using a classifier on pictoral data: while the data is presented ``graphically", the participant is actually classifying the data based on underlying summary statistics.}
\comment{that last sentance is probably reaching a bit far...}
% In figure~\ref{fig:visualaptitudects}, we see that participant performance on lineups is positively correlated with performance on card rotation, figure classification, and paper folding tasks. This suggests that skills associated with visual reasoning ability are related to lineup performance. As participants must use the same skills in lineups (mental rotation, classification and determining categorization schemes, and multi-step spatial reasoning) as in the factor-referenced tests, this is not particularly surprising. In addition,there seems to be some positive relationship between a participant's score on the visual search task and their score on lineups: the visual search task represents a baseline of a participant's ability to find a matching pattern, while lineups require that task as well as the ability to determine what the pattern is for a particular graph. Even excluding the one low visual search score that is a high-leverage point, there seems to be a positive relationship between a participant's score on lineups and their score for visual search. 
% 
% <<VisReasoningCts,echo=FALSE, include=FALSE, eval=T, fig.width=8,fig.height=8, message=F, warning=F>>=
% lineup.summary$variable <- factor(lineup.summary$variable, labels=c("Card Rotation", "figure classification", "Paper Folding", "Visual Search"))
%  
% qplot(data=lineup.summary, x=value, y=lineup, geom="point", shape=I(1), size=I(3)) + facet_wrap(~variable, scales="free_x") + xlab("Score") + ylab("Lineup Score") + geom_smooth(method="lm") + theme_bw()
% @

% 
% \begin{figure}[hbt]\centering
% \includegraphics[width=0.75\linewidth]{fig-VisReasoningCts-1}
% \caption[Visual Aptitude Study Results]{Scatterplots of all test scores compared to participants' scores in the lineup tests. There is a relatively strong positive correlation between lineup score and scores on visuospatial reasoning tests.}\label{fig:visualaptitudects}
% \end{figure}

@@

% \section{Lineup Performance: Accounting for ``Visual Intelligence''}
% As figure~\ref{fig:biplots}(a) shows, the first principal component seems to primarily describe ``general intelligence", or ``visual intelligence" in a generic sense (as we have only tested visual skills). As the visual search task (VST) is the most generic of visual tasks (find the matching pattern) and requires no spatial manipulation, rotation, or inference, we might use it as an indicator of general visual intelligence (or cognitive tempo and attention to detail). 
% 
% <<partialregression, results='asis', echo=1, tidy=T>>=
% model1 <- lm(lineup ~ vis_search, data=ans.summary)
% print(xtable(summary(model1), caption="Results of the linear regression of visual search score on lineup score. \\label{vst-reg}."), sanitize.text.function=sanitize)
% @
% <<partialregression1,fig.keep='all', fig.width=4, fig.height=4, out.width='.45\\linewidth', fig.cap="Biplots of the adjusted lineup, card rotation, figure classification, and paper folding scores principal components analysis. Adjusted lineup scores are most associated with figure classification in the second principal component. In the third and fourth components, adjusted scores are almost entirely orthogonal. \\label{fig:biplot-adjusted}", tidy=T, echo=1:7>>=
% # partial regression data set
% ans.part <- ans.summary
% ans.part$lineup <- resid(lm(lineup ~ vis_search, data=ans.summary))
% ans.part$card_rot <- resid(lm(card_rot ~ vis_search, data=ans.summary))
% ans.part$fig_class <- resid(lm(fig_class ~ vis_search, data=ans.summary))
% ans.part$folding <- resid(lm(folding ~ vis_search, data=ans.summary))
% 
% pca2 <- prcomp(ans.part[,c("lineup", "card_rot", "fig_class", "folding")], scale=T, retx=T)
% biplot(pca2, choices=1:2, pc.biplot=T, cex=c(.5, 1), adj=.75)
% biplot(pca2, choices=3:4, pc.biplot=T, cex=c(.5, 1), adj=.75)
% @
% <<partialregression2, echo=F, results='asis'>>=
% model2 <- lm(lineup~fig_class+card_rot+folding, data=ans.part)
% print(xtable(model2, caption="Results of the linear regression of (the ETS tests $|$ VST) on (lineup score $|$ VST)"), sanitize.text.function=sanitize, caption.placement="top")
% @
% Removing paper folding from the analysis due to its' obvious nonsignificance and collinearity with card rotation, we get a slightly better regression result. 
% <<partialregression3, echo=F, results='asis'>>=
% model3 <- lm(lineup~fig_class+card_rot, data=ans.part)
% print(xtable(model3, caption="Results of the reduced linear regression of (the ETS tests $|$ VST) on (lineup score $|$ VST)"), sanitize.text.function=sanitize, caption.placement="top")
% print(xtable(anova(model3, model2), caption="Model comparison between full model (including paper folding) and reduced model (excluding paper folding), when accounting for scores on the VST in both independent and dependent variables. "), sanitize.text.function=sanitize, caption.placement="top")
% @
% 
% Comparing the full and reduced models, we see that the reduction does not significantly reduce the explained sum of squares. 
% 
% 
% Even using VST as a ``general intelligence" factor, the first PC still does not discriminate between the tests. This approach is probably not worth pursuing further. 
<<biplots-pca5, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.8\\linewidth', comment="|", fig.cap="Plots of principal components 2-5 with observations. \\label{fig:biplots5}", echo=F, fig.env="figure">>=
par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
biplot(pca, choices=2:3, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
biplot(pca, choices=4:5, pc.biplot=T, cex=c(.5, 1), xlim=c(-2,2), ylim=c(-2,2))
@

@@

# use AIC in backward selection
library(MASS)
m1 <- stepAIC(m0, direction="both")
@
<<modelresults>>=
anova(m1)
summary(m1)
@
\end{document}

>>>>

  rbind.fill(
    data.frame(
      xvar = names(data)[ycol], 
      yvar = names(data)[xcol],
      x = data[, xcol], 
      y = data[, ycol], 
      points = T,
      data
    ),
    data.frame(
      yvar = names(data)[ycol], 
      xvar = names(data)[xcol],
      x = mean(range(data[, ycol])), 
      y = mean(range(data[, xcol])),
      label = round(cor(data[,xcol], data[,ycol]), 3),
      points = F
    )
  )
}))
all$xvar <- factor(all$xvar, levels=names(data))
all$yvar <- factor(all$yvar, levels=names(data))
densities <- do.call("rbind", lapply(1:ncol(data), function(i) {
    tmp <- density(data[,i], cut = 1, adjust=.75)
    data.frame(
      xvar = names(data)[i], 
      yvar = names(data)[i],
      x = tmp$x, 
      y = tmp$y/max(tmp$y)* diff(range(tmp$x)) + min(tmp$x)
    )
  }))
fix.names <- function(values){
  values %>% 
    str_replace("lineup", "Lineups") %>% 
    str_replace("vis_search", "Visual Search") %>%
    str_replace("fig_class", "Figure Classification") %>% 
    str_replace("card_rot", "Card Rotation") %>%
    str_replace("folding", "Paper Folding")
}

@@

\newtext{This emphasizes the underpinnings of lineups: they are a visual medium, but they ultimately are a classification task, presented in a graphical manner. Using lineups as a proxy for statistical significance tests is similar to using a classifier on pictoral data: while the data is presented ``graphically", the participant is actually classifying the data based on underlying summary statistics.}
\comment{that last sentance is probably reaching a bit far...}

@@

<<biplots-pca5, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.8\\linewidth', comment="|", fig.cap="Plots of principal components 2-5 with observations. \\label{fig:biplots5}", echo=F, fig.env="figure">>=
par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
biplot(pca, choices=2:3, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
biplot(pca, choices=4:5, pc.biplot=T, cex=c(.5, 1), xlim=c(-2,2), ylim=c(-2,2))
@

# use AIC in backward selection
library(MASS)
m1 <- stepAIC(m0, direction="both")
@
<<modelresults>>=
anova(m1)
summary(m1)
@

\end{document}

@@
